Bayesian inference of relative fitness on high-throughput pooled competition assays

### 2023-11-20 doc/_variational_inference_primer.qmd
remove repeated label

Variational inference circumvents these technical challenges by proposing an
approximate solution to the problem. Instead of working with the posterior
distribution in its full glory $\pi(\underline{\theta} \mid \underline{x})$, let
us propose an approximate posterior distribution $q_\phi$ that belongs to a
distribution family fully parametrized by $\phi$. For example, let us say that
the distribution $q_\phi$ belongs to the family of multivariate Normal
distributions such that $\phi = (\underline{\mu},
\underline{\underline{\Sigma}})$, where $\underline{\mu}$ is the vector of means
and $\underline{\underline{\Sigma}}$ is the covariance matrix. If we replace
$\pi$ by $q_\phi$, we want $q_\phi$ to resemble the original posterior as much
as possible. Mathematically, this can be expressed as minimizing a "*distance
metric*"---the Kullback-Leibler (KL) divergence, for example---between the
distributions. Note that we use quotation marks because, formally, the KL
divergence is not a distance metric since it is not symmetric. Nevertheless, the
variational objective is set to find a distribution $q_\phi^*$ such that
$$
q_\phi^*(\underline{\theta}) =
\min_\phi D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right),
$${#eq-vi_objective}
where $D_{KL}$ is the KL divergence. Furthermore, we highlight that the KL
divergence is a strictly positive number, i.e.,
$$
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) \geq 0,
$${#eq-vi_kl_pos}
as this property will become important later on.

At first sight, @eq-vi_objective does not improve the situation but only
introduces further technical complications. After all, the definition of the KL
divergence
$$
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) \equiv 
\int \cdots \int d^K\underline{\theta}\;
q_\phi(\underline{\theta})
\ln \frac{
    q_\phi(\underline{\theta})
}{
    \pi(\underline{\theta} \mid \underline{x})
},
$${#eq-vi_kl_div}
includes the posterior distribution $\pi(\underline{\theta} \mid \underline{x})$
we are trying to get around. However, let us manipulate @eq-vi_kl_div to beat it
to a more reasonable form. First, we can use the properties of the logarithms to
write
$$
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) = 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) -
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{\theta} \mid \underline{x}),
$${#eq-vi_step01}
where, for convenience, we write a single integration sign
($d^K\underline{\theta}\;$ still represents a multi-dimensional differential). For
the second term in @eq-vi_step01, we can substitute the term inside the
logarithm using @eq-vi_bayes_theorem. This results in
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) \\
&- \int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \left( 
    \frac{
        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
    }{
        \pi(\underline{x})
    }
\right).
\end{aligned}
$${#eq-vi_step02}
Again, using the properties of logarithms, we can split @eq-vi_step02, obtaining
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) \\
&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{\theta}) \\
&+\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{x}).
\end{aligned}
$${#eq-vi_step03}
It is convenient to write @eq-vi_step03 as
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \frac{
    q_\phi(\underline{\theta})
    }{
        \pi(\underline{\theta})
    } \\
&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&+ \ln \pi(\underline{x}) 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta}),
\end{aligned}
$${#eq-vi_step04}
where for the last term, we can take $\ln \pi(\underline{x})$ out of the
integral since it does not depend on $\underline{\theta}$. Lastly, we utilize
two properties:

>>>>

Variational inference circumvents these technical challenges by proposing an
approximate solution to the problem. Instead of working with the posterior
distribution in its full glory $\pi(\underline{\theta} \mid \underline{x})$, let
us propose an approximate posterior distribution $q_\phi$ that belongs to a
distribution family fully parametrized by $\phi$. For example, let us say that
the distribution $q_\phi$ belongs to the family of multivariate Normal
distributions such that $\phi = (\underline{\mu},
\underline{\underline{\Sigma}})$, where $\underline{\mu}$ is the vector of means
and $\underline{\underline{\Sigma}}$ is the covariance matrix. If we replace
$\pi$ by $q_\phi$, we want $q_\phi$ to resemble the original posterior as much
as possible. Mathematically, this can be expressed as minimizing a "*distance
metric*"---the Kullback-Leibler (KL) divergence, for example---between the
distributions. Note that we use quotation marks because, formally, the KL
divergence is not a distance metric since it is not symmetric. Nevertheless, the
variational objective is set to find a distribution $q_\phi^*$ such that
$$
q_\phi^*(\underline{\theta}) =
\min_\phi D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right),
$${#eq-vi_objective_si}
where $D_{KL}$ is the KL divergence. Furthermore, we highlight that the KL
divergence is a strictly positive number, i.e.,
$$
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) \geq 0,
$${#eq-vi_kl_pos}
as this property will become important later on.

At first sight, @eq-vi_objective_si does not improve the situation but only
introduces further technical complications. After all, the definition of the KL
divergence
$$
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) \equiv 
\int \cdots \int d^K\underline{\theta}\;
q_\phi(\underline{\theta})
\ln \frac{
    q_\phi(\underline{\theta})
}{
    \pi(\underline{\theta} \mid \underline{x})
},
$${#eq-vi_kl_div}
includes the posterior distribution $\pi(\underline{\theta} \mid \underline{x})$
we are trying to get around. However, let us manipulate @eq-vi_kl_div to beat it
to a more reasonable form. First, we can use the properties of the logarithms to
write
$$
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) = 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) -
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{\theta} \mid \underline{x}),
$${#eq-vi_step01}
where, for convenience, we write a single integration sign
($d^K\underline{\theta}\;$ still represents a multi-dimensional differential). For
the second term in @eq-vi_step01, we can substitute the term inside the
logarithm using @eq-vi_bayes_theorem. This results in
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) \\
&- \int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \left( 
    \frac{
        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
    }{
        \pi(\underline{x})
    }
\right).
\end{aligned}
$${#eq-vi_step02}
Again, using the properties of logarithms, we can split @eq-vi_step02, obtaining
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) \\
&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{\theta}) \\
&+\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{x}).
\end{aligned}
$${#eq-vi_step03}
It is convenient to write @eq-vi_step03 as
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \frac{
    q_\phi(\underline{\theta})
    }{
        \pi(\underline{\theta})
    } \\
&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&+ \ln \pi(\underline{x}) 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta}),
\end{aligned}
$${#eq-vi_step04}
where for the last term, we can take $\ln \pi(\underline{x})$ out of the
integral since it does not depend on $\underline{\theta}$. Lastly, we utilize
two properties:
